---
title: "That was close! A counterfactual simulation model of causal judgments about decisions"
author: "Sarah A. Wu, Shruti Sridhar, and Tobias Gerstenberg"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  bookdown::html_document2:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    highlight: tango
---

# Setup

```{r setup, include = F, message = F, warning = F}
# load relevant libraries and functions
require(knitr)         # for knitting        
library(ggstatsplot)   # extension of ggplot2
library(lme4)          # for mixed effects models
library(lmerTest)      # for p-values
library(DT)            # for datatable() function 
library(Metrics)       # for RMSE
library(Hmisc)         # for bootstrapped confidence intervals
library(ggrepel)       # avoid overlap in geom_text
library(patchwork)     # for multiple plots
library(tidyverse)     # for everything else

knitr::opts_chunk$set(echo = T, warning = F, message = F)

# set default plot theme 
theme_set(theme_classic() + 
            theme(text = element_text(size = 14))) 

# suppress warnings about grouping 
options(dplyr.summarise.inform = F)

options(xtable.floating = F)
options(xtable.timestamp = "")

data_dir = '../../data/'
figures_dir = '../../figures/'
trial_info_dir = '../experiment/'
```

# Data

## Trial info

```{r importing_trial_info}
trial_info = read_csv(paste0(trial_info_dir, 'experiment.csv'),
                      show_col_types = F) %>%
  mutate(bar_color = ifelse(cf_success_rate >= 50, 1, 0),
         pred_cf = ifelse(outcome == 'won', 100 - cf_success_rate, cf_success_rate),
         pred_hyp = ifelse(outcome == 'won', 100 - hyp_success_rate, hyp_success_rate),
         diff_made = c(F, F, T, T, T, T,
                       F, T, T, F, T, F,
                       T, T, F, F, F, F))
```

## Counterfactual condition

### Importing

```{r importing_cf}
data_cf_raw = read_csv(paste0(data_dir, 'experiment1_cf/trials.csv'),
                       show_col_types = F) %>%
  rename(id = workerid)

data_cf_participants_raw = read_csv(paste0(data_dir, 'experiment1_cf/participants.csv'),
                                    show_col_types = F) %>%
  rename(id = workerid)
```

### Processing

```{r processing_cf}
data_cf = data_cf_raw %>%
  drop_na(trial) %>%
  left_join(trial_info %>%
              select(trial, outcome),
            by = c('trial')) %>%
  mutate(cf_recode = ifelse(outcome == 'won', 100 - cf, cf),
         id = factor(id))

data_cf_means = data_cf %>%
  group_by(trial) %>%
  summarise(cf_mean = mean(cf),
            cf_low = smean.cl.boot(cf)[2],
            cf_high = smean.cl.boot(cf)[3],
            cf_recode_mean = mean(cf_recode),
            cf_recode_low = smean.cl.boot(cf_recode)[2],
            cf_recode_high = smean.cl.boot(cf_recode)[3])

data_cf_participants = data_cf_participants_raw %>%
  mutate(id = factor(id))
```


Number of participants:
```{r n_participants_cf}
length(unique(data_cf$id))
```

### Feedback

Participants were asked: "What factors influenced how you decided to respond? Do you have any questions or comments regarding the experiment?"

```{r feedback_cf}
data_cf_participants %>%
  select(feedback) %>% 
  datatable()
```


## Hypothetical condition

### Importing

```{r importing_hyp}
data_hyp_raw = read_csv(paste0(data_dir, 'experiment1_hyp/trials.csv'),
                       show_col_types = F) %>%
  rename(id = workerid)

data_hyp_participants_raw = read_csv(paste0(data_dir, 'experiment1_hyp/participants.csv'),
                                    show_col_types = F) %>%
  rename(id = workerid)
```

### Processing

```{r processing_hyp}
data_hyp = data_hyp_raw %>%
  drop_na(trial) %>%
  left_join(trial_info %>%
              select(trial, outcome),
            by = c('trial')) %>%
  mutate(hyp_recode = ifelse(outcome == 'won', 100 - hyp, hyp),
         id = factor(id))

data_hyp_means = data_hyp %>%
  group_by(trial) %>%
  summarise(hyp_mean = mean(hyp),
            hyp_low = smean.cl.boot(hyp)[2],
            hyp_high = smean.cl.boot(hyp)[3],
            hyp_recode_mean = mean(hyp_recode),
            hyp_recode_low = smean.cl.boot(hyp_recode)[2],
            hyp_recode_high = smean.cl.boot(hyp_recode)[3])

data_hyp_participants = data_hyp_participants_raw %>%
  mutate(id = factor(id))
```


Number of participants:
```{r n_participants_hyp}
length(unique(data_hyp$id))
```

### Feedback

Participants were asked: "What factors influenced how you decided to respond? Do you have any questions or comments regarding the experiment?"

```{r feedback_hyp}
data_hyp_participants %>%
  select(feedback) %>% 
  datatable()
```


## Causal condition

### Importing

```{r importing_cause}
data_raw = read_csv(paste0(data_dir, 'experiment2/trials.csv'),
                    show_col_types = F) %>%
  rename(id = workerid)

data_participants_raw = read_csv(paste0(data_dir, 'experiment2/participants.csv'),
                                 show_col_types = F) %>%
  rename(id = workerid)
```

### Processing

```{r processing_cause}
data = data_raw %>%
  drop_na(trial) %>%
  left_join(trial_info, by = c('trial')) %>%
  mutate(id = factor(id))

data_participants = data_participants_raw %>%
  mutate(id = factor(id))
```


Number of participants:
```{r n_participants_cause}
length(unique(data$id))
```

### Feedback

Participants were asked: "What factors influenced how you decided to respond? Do you have any questions or comments regarding the experiment?"

```{r feedback_cause}
data_participants %>%
  select(feedback) %>% 
  datatable()
```




# Demographics

## Helper function
```{r demographics_helper}
print_demographics = function(data) {
  # gender
  data %>%
    group_by(gender) %>%
    summarise(n = n()) %>%
    print()
  
  # age
  print('age:')
  mean(data$age, na.rm = T) %>% print()
  sd(data$age, na.rm = T) %>% print()
  
  # race
  data %>%
    group_by(race) %>%
    summarise(n = n()) %>%
    print()
  
  # ethnicity
  data %>%
    group_by(ethnicity) %>%
    summarise(n = n()) %>%
    print()

  # time taken
  print('time taken (min):')
  print(mean(data$time, na.rm = T)/60/1000)
  print(sd(data$time, na.rm = T)/60/1000)
}
```


## Experiment 1

```{r demographics_exp1}
print_demographics(data_cf_participants %>%
                     rbind(data_hyp_participants))
```

## Experiment 2

```{r demographics_exp2}
print_demographics(data_participants)
```


# Stats and modeling

## Importing features

```{r importing_features}
features = read_csv(sprintf('%s/features.csv', trial_info_dir),
                    show_col_types = F) %>%
  left_join(trial_info %>%
              select(trial, outcome),
            by = 'trial') %>%
  mutate(doors_final = factor(doors_final,
                              levels = c('open_open', 'open_closed',
                                         'closed_open', 'closed_closed',
                                         'none_none')))

data_modeling = data %>%
  select(id, trial, cause) %>%
  left_join(data_cf_means %>%
              select(trial, cf_recode_mean),
            by = 'trial') %>%
  left_join(data_hyp_means %>%
              select(trial, hyp_recode_mean),
            by = 'trial') %>%
  left_join(features,
            by = 'trial')
```

## Fitting all trials

```{r fit_heuristic, message = T}
fit.heuristic = lmer(
  formula = cause ~ 1 + outcome:doors_final + (1 | id),
  data = data_modeling
)

summary(fit.heuristic)
```
Fit matrix is rank-deficient because not all factor levels have data for both outcomes (e.g. there is no trial that was a loss and had no doors). So we fit win and loss trials separately, using the same formula.


## Fitting wins only

```{r fit_heuristic_win}
fit.heuristic_win = lmer(
  formula = cause ~ 1 + doors_final + (1 | id),
  data = data_modeling %>%
    filter(outcome == 'won')
)

summary(fit.heuristic_win)
```

## Fitting losses only

```{r fit_heuristic_loss}
fit.heuristic_loss = lmer(
  formula = cause ~ 1 + doors_final + (1 | id),
  data = data_modeling %>%
    filter(outcome == 'lost')
)

summary(fit.heuristic_loss)
```


# Plots

## Experiment 1

### Set up data
```{r exp1_plot_data}
data_cf_pred = data_cf_means %>%
  rename(emp_mean = cf_mean,
         emp_low = cf_low,
         emp_high = cf_high) %>%
  left_join(trial_info %>%
          select(trial,
                 pred = cf_success_rate,
                 pred_recode = pred_cf,
                 diff_made),
        by = 'trial')

data_hyp_pred = data_hyp_means %>%
  rename(emp_mean = hyp_mean,
         emp_low = hyp_low,
         emp_high = hyp_high) %>%
  left_join(trial_info %>%
          select(trial,
                 pred = hyp_success_rate,
                 pred_recode = pred_hyp,
                 diff_made),
        by = 'trial')
```


### Helper function
```{r exp1_plot_helper}
limits = c(0, 100)
color_scale = c("FALSE" = "red3", "TRUE" = "chartreuse3")

plot_simulation = function(data, name) {
  r = cor(data$emp_mean, data$pred) %>%
    round(2)
  rmse = rmse(data$emp_mean, data$pred)
  
  g = ggplot(data = data,
             aes(x = pred,
                 y = emp_mean)) +
    geom_abline(intercept = 0, slope = 1,
                linetype = 2, size = 0.2) +
    # error bars
    geom_linerange(size = 0.5, 
                   mapping = aes(ymin = emp_low,
                                 ymax = emp_high),
                   color = 'gray') +
    # means
    geom_point(mapping = aes(fill = 'darkgray'),
               shape = 21, size = 3, stroke = 0.2) +
    # trial labels
    geom_text_repel(aes(label = trial),
                    size = 4,
                    seed = 1,
                    max.overlaps = 18) +
    theme(legend.position = "none",
          text = element_text(size = 12),
          axis.text = element_text(size = 12)) +
    annotate('text',
             label = sprintf('RMSE = %.2f\nr = %s', rmse, r),
             hjust = 0,   # left align
             x = limits[1], y = limits[2]) +
    scale_x_continuous(name = paste(name, 'simulation'),
                       limits = limits) +
    scale_y_continuous(name = paste(name, 'judgment'),
                       limits = limits,
                       expand = expansion(0.1)) +
    scale_color_manual(values = color_scale,
                      guide = 'none') +
    scale_fill_manual(values = color_scale,
                      guide = 'none')

  return(g)
}
```

### Scatterplots

```{r exp1_plot, fig.width = 6, fig.height = 3}
p_cf = plot_simulation(data_cf_pred, 'counterfactual')
p_hyp = plot_simulation(data_hyp_pred, 'hypothetical')

p_hyp + p_cf + plot_annotation(tag_levels = 'A')

#ggsave(paste0(figures_dir, 'experiment1.pdf'), width = 6, height = 3)
```


## Experiment 2

### Set up data
```{r exp2_plot_data}
data_pred = data %>%
  group_by(trial) %>%
  summarise(cause_mean = mean(cause),
            cause_low = smean.cl.boot(cause)[2],
            cause_high = smean.cl.boot(cause)[3]) %>%
  ungroup() %>%
  # counterfactual judgments
  left_join(data_cf_means %>%
              select(trial,
                     pred_cf_mean = cf_recode_mean,
                     pred_cf_low = cf_recode_low,
                     pred_cf_high = cf_recode_high),
            by = 'trial') %>%
  # hypothetical judgments
  left_join(data_hyp_means %>%
              select(trial,
                     pred_hyp_mean = hyp_recode_mean,
                     pred_hyp_low = hyp_recode_low,
                     pred_hyp_high = hyp_recode_high),
            by = 'trial') %>%
  # heuristic model
  left_join(features %>%
              select(trial, outcome),
            by = 'trial') %>%
  left_join(data_modeling %>%
              filter(outcome == 'won') %>%
              mutate(x = predict(object = fit.heuristic_win,
                                 newdata = data_modeling %>%
                                   filter(outcome == 'won'),
                                 re.form = NA)) %>%
              rbind(data_modeling %>%
                      filter(outcome == 'lost') %>%
                      mutate(x = predict(object = fit.heuristic_loss,
                                         newdata = data_modeling %>%
                                           filter(outcome == 'lost'),
                                         re.form = NA))) %>%
              group_by(trial) %>%
              summarise(pred_heuristic_mean = mean(x)),
            by = 'trial') %>%
  mutate(# since the plot function will draw error bars
         pred_heuristic_low = pred_heuristic_mean,
         pred_heuristic_high = pred_heuristic_mean) %>%
  merge(trial_info %>%
          select(trial, diff_made),
        by = c('trial'))
```


### Helper function
```{r exp2_plot_helper}
limits = c(0, 100)
color_scale = c("FALSE" = "red3", "TRUE" = "chartreuse3")

plot_cause_vs = function(model_name) {
  col_mean = paste('pred', model_name, 'mean', sep = '_')
  col_low = paste('pred', model_name, 'low', sep = '_')
  col_high = paste('pred', model_name, 'high', sep = '_')
  
  r = cor(data_pred$cause_mean, data_pred[[col_mean]]) %>%
    round(2)
  rmse = rmse(data_pred$cause_mean, data_pred[[col_mean]])

  g = ggplot(data = data_pred,
             aes(x = get(col_mean),
                 y = cause_mean)) +
    geom_abline(intercept = 0, slope = 1,
                linetype = 2, size = 0.2) +
    # error bars
    geom_linerange(size = 0.5,
                   mapping = aes(ymin = cause_low,
                                 ymax = cause_high),
                   color = 'gray') +
    geom_errorbarh(mapping = aes(xmin = get(col_low),
                               xmax = get(col_high)),
                   color = 'lightgray',
                   height = 0) +
    # means
    geom_point(mapping = aes(fill = diff_made),
               shape = 21, size = 3, stroke = 0.2) +
    # trial labels
    geom_text_repel(aes(label = trial,
                        color = diff_made),
                    size = 4,
                    seed = 1,
                    max.overlaps = 18) +
    theme(legend.position = "none",
          text = element_text(size = 12),
          axis.text = element_text(size = 12)) +
    annotate('text',
             label = sprintf('RMSE = %.2f\nr = %s', rmse, r),
             hjust = 0,
             x = limits[1], y = limits[2] - 3) +
    scale_x_continuous(name = case_when(
      model_name == 'cf' ~ 'counterfactual simulation model',
      model_name == 'hyp' ~ 'hypothetical simulation model',
      model_name == 'heuristic' ~ 'heuristic model'
    ),
                       limits = limits) +
    scale_y_continuous(name = 'causal judgment',
                       limits = limits) +
    scale_color_manual(values = color_scale,
                       guide = 'none') +
    scale_fill_manual(values = color_scale,
                      guide = 'none')
  
  return(g)
}
```

### Scatterplots

```{r exp2_plot, fig.width=10, fig.height=3.5}
p_cause_vs_cf = plot_cause_vs('cf')
p_cause_vs_hyp = plot_cause_vs('hyp')
p_cause_vs_heuristic = plot_cause_vs('heuristic')

p_cause_vs_hyp + p_cause_vs_cf + p_cause_vs_heuristic +
  plot_annotation(tag_levels = 'A')

#ggsave(paste0(figures_dir, 'experiment2.pdf'), width = 10, height = 3.5)
```

# Session info

```{r}
sessionInfo()
```

